The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-05-02 Paper.Rmd
edits, mostly for new holder materials
various edits, getting ready for submission

---
title: "The importance of a complete statistical description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, Philip C. Reid, Daniel C. Reuman"
date: ""
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

@@

Datasets included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data. Copula structure describing 
the relationship between atmospheric weather variables such as rainfall or wind speed 
as measured in multiple locations has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [[RaCA](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/soils/survey/?cid=nrcs142p2_054164); @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for sampling locations).
Measurments were from the central pedon of a soil core, to 1 m depth.<!--DAN: we should ask Terry to explain pedon/ R horizon-->
<!--***DAN: Terry to check the wording and accuracy here-->
Species-trait data were average species basal metabolic rate (KJ per hour) and 
body mass (grams) data for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years; these are 
indicators of aphid migration counts. Data pre-processing for the aphid data was the same as 
that of @Sheppard2016,
but we excluded one of the 11 sites they used in order to only use sites with at least 30 
years data.  Our second population dataset comprised average annual abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 plankton taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. These data 
were obtained from the Continuous Plankton Recorder (CPR) dataset now operated and maintained by 
the Marine Biological Association. Data pre-processing was similar but not identical to that 
of @Sheppard2017; see Appendix \ref{SM-Data_details} for details.
Community-level data obtained from the Cedar Creek Ecosystem Science Reserve were <!--DAN : check this rephrasing-->
plant aboveground biomass and plant percent cover data (we used this percent cover data to calculate 
Shannon's diversity index) for `r nrow(d_cc[[5]])` plots, each 9m by 9m, as sampled in 
the year 2000 [@Tilman2001; @Tilman2006] <!--actually they collected data in each year of 1996,1997,1998,1999,2000,2007-->. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. 
Sampling locations spanned a hydrologic gradient from upland to seasonal wetland.
See Appendix \ref{SM-Data_details} for additional data details.

The environmental, species-trait, and community datasets are henceforth 
called the "bivariate" datasets in the sense that they comprise two quantities measured at 
different locations or for different species. The phenology and population datasets are 
"multivariate" in that they comprise, for each taxon, measurements through time at multiple locations; 
copula structure can be studied for each pair of locations. Likewise the ecosystem 
functioning (methane flux) dataset was  multivariate in this sense (see \ref{tab_data_info}).

<!--summary table with info of all data used in Paper-->
```{r tab_summary_data_info, echo=F, results="asis",message=F}
library(kableExtra)
library(dplyr)
tab_data_info<-as.data.frame(matrix(NA,nrow=8,ncol=4))
colnames(tab_data_info)<-c("Data","Category","No. of measurements we used", "References")
tab_data_info$Data<-c("Soil C and N","Bird masses and BMR", "Mammal masses and BMR", "Cedar Creek data", 
                      "Green spruce aphid abundance", "Leaf-curling plum aphid first flight","Ceratium furca abundance","Methane flux")
tab_data_info$Category<-c("Environmental","Species-trait","Species-trait","Community level",
                          "Population","Phenological","Population","Ecosystem-functioning")
tab_data_info$`No. of measurements we used`<-c(paste(nrow(d_soilCN)," locations",sep=""),
                                               paste(nrow(d_birdsBMR)," birds",sep=""),
                                               paste(nrow(d_mammalsBMR)," mammals",sep=""),
                                               paste(nrow(d_cc[[5]])," plots",sep=""),
                                               "10 locations with atleast 30 yrs. of timeseries for each 20 sp.",
                                               "10 locations with atleast 30 yrs. of timeseries for each 20 sp.",
                                               "14 locations with atleast 45 yrs. of timeseries for each 22 taxa",
                                               "13 locations with atleast 50 dates of data")
tab_data_info$References<-c("US Rapid Carbon Assessment database (RaCA), Wills et al. (2014)", "McNab 2009","McNab 2008",
                            "Cedar Creek Ecosystem Science Reserve [biodiversity experiment e120]",
                            "Rothamsted Insect Survey","Rothamsted Insect Survey","Continuous Plankton Recorder Survey",
                            "Great Miami Wetland Mitigation Bank, Smyth et al. (2019)")

knitr::kable(tab_data_info, 
             format="latex", align="l",
             caption = "Summary table for data we used, bolds are the multivariate data, rest are bivariates. \\label{tab_data_info}", 
             booktabs = T,linesep = "\\addlinespace") %>%
             column_spec(1:4,width="2.8 cm")%>%
             row_spec(5:8, bold=T)
```

@@

The third mechanism we explored is evolutionary, and pertains to bivariate 
trait data across species, e.g., to our bird and mammal BMR versus body mass 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
We simulated bivariate character evolution on an estimate of 
phylogeny for $817$ mammal species taken from @GenoudIM2018.
The root character state and the change 
across each branch were randomly chosen from matrices of one million different
bivariate draws from distributions showing one of five distinct types of copula 
structure: 1) extreme left-tail dependence, 2) moderate left-tail dependence,
3) symmetric tail dependence, 4) moderate right-tail dependence, and 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech} for details). 
All marginal distributions were standard normal, and distributions all had Spearman 
correlation $0.875$, so simulations assess the impact of copula structure of character
changes on evolutionary outcomes, controlling for marginals and correlations of character
changes. For each of the five copula structures, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters was assessed across the phylogeny tips.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

<!--Holder stuff-->
Our simulations of character evolution suggest the hypothesis that evolutionary
changes in bird and mammal BMR and body size may exhibit right-tail dependence. This is
a hypothesis only, because the tail dependence of Fig. \ref{fig_biv_multi_raw_cop}F,G
could have come about in another, unknown way: our simulations show that tail dependence 
in evolutionary changes is sufficient, but may not be necessary, to produce tail
dependence in characters of extant species. Even if the hypothesized mechanism
is correct, our results simply replace one question, i.e., why do we see right-tail
dependence between BMR and body mass for birds and mammals, with another question,
i.e., why might we see right-tail dependence in evolutionary changes of these traits.
See Appendix \ref{SM-evol_mech} for additional thoughts on the potential importance of 
copulas for studies of character evolution when phylogeny branch lengths are not available.

<!-- We have only given three mechanisms, but there are likely to be others, and discoverig
additional mechanisms of copula structure is certain worthy of future work. -->
Other mechanisms that we have not considered likely also operate to produce copula structure
in ecological and evolutionary datasets, and should be considered in future work. For instance, 
measurement error may modify copula structure, though one would have to have information
on the precise nature of measurement error in a dataset to explore its influence. 
Not only did the models we used to explore mechanisms not consider measurement error, 
they were extremely simple in other respects, as well, not including phenomena such as 
delayed density dependence, dispersal, stage structure in population dynamics, and trophic
interactions.
We hope by enumerating three potential mechanisms we help inspire additional research on the
potentially numerous mechanisms of copula structure that may apply in diverse datasets.

@@

We thank the many contributors to the large datasets we used; D. Stevens and P. Verrier for data 
extraction; and Joel E. Cohen, David Tilman, Lauren Hallet, Jim Bever, Jonathan Walter, Thomas 
Anderson, and Lei Zhao for helpful suggestions. We thank James Bell of the Rothamsted Insect Survey 
(RIS). 
The RIS, a UK Capability, is funded by the Biotechnology and Biological Sciences Research Council 
under the Core Capability Grant BBS/E/C/000J0200. SG, LWS and DCR were partly funded by US National 
Science Foundation grants 0075516 and 1442595 and the James S McDonnell Foundation.

>>>>

---
title: "The importance of a complete statistical description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, Philip C. Reid, James D. Bever, Daniel C. Reuman"
date: ""
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

Bever: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

@@

Datasets included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data (Table \ref{tab_data_info}). Copula structure describing 
the relationship between atmospheric weather variables such as rainfall or wind speed 
as measured in multiple locations has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [RaCA; @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for sampling locations).
Measurments were from the central pedon of a soil core, to 1 m depth.
Species-trait data were average species basal metabolic rate (BMR, KJ per hour) and 
body mass (grams) data for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years; these are 
indicators of aphid migration counts. Data pre-processing for the aphid data was the same as 
that of @Sheppard2016,
but we excluded one of the 11 sites they used in order to only use sites with at least 30 
years data.  Our second population dataset comprised average annual plankton abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. These data 
were obtained from the Continuous Plankton Recorder (CPR) dataset, now operated and maintained by 
the Marine Biological Association of the United Kingdom. Data pre-processing was similar but not identical to that 
of @Sheppard2017; see Appendix \ref{SM-Data_details} for details.
Community-level data, obtained from the Cedar Creek Ecosystem Science Reserve, were 
plant aboveground biomass and Shannon's diversity index for `r nrow(d_cc[[5]])` plots, each 9m by 9m, as sampled in 
the years 1996-2000 and 2007, each year analyzed separately [@Tilman2001; @Tilman2006]. 
Shannon's index was computed from plant species percent cover data.  
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. 
Sampling locations spanned a hydrologic gradient from upland to seasonal wetland.
See Appendix \ref{SM-Data_details} for additional data details.

The environmental, species-trait, and community datasets are henceforth 
called the "bivariate" datasets in the sense that they comprise two quantities measured at 
different locations or for different species. The phenology and population datasets are 
"multivariate" in that they comprise, for each taxon, measurements through time at multiple locations; 
copula structure can be studied for each pair of locations. Likewise the ecosystem 
functioning (methane flux) dataset was  multivariate in this sense (Table \ref{tab_data_info}).

<!--summary table with info of all data used in Paper-->
```{r tab_summary_data_info, echo=F, results="asis",message=F}
library(kableExtra)
library(dplyr)
tab_data_info<-as.data.frame(matrix(NA,nrow=8,ncol=4))
colnames(tab_data_info)<-c("Data","Category","No. of measurements we used", "References")
tab_data_info$Data<-c("Soil C and N","Bird body masses and BMR", "Mammal body masses and BMR", "Cedar Creek data", 
                      "Green spruce and other aphid species abundances", "Leaf-curling plum and other aphid first flight dates","Ceratium furca and other plankton taxa abundances","Methane flux")
tab_data_info$Category<-c("Environmental","Species-trait","Species-trait","Community level",
                          "Population","Phenological","Population","Ecosystem functioning")
tab_data_info$`No. of measurements we used`<-c(paste(nrow(d_soilCN)," locations",sep=""),
                                               paste(nrow(d_birdsBMR)," birds",sep=""),
                                               paste(nrow(d_mammalsBMR)," mammals",sep=""),
                                               paste(nrow(d_cc[[5]])," plots",sep=""),
                                               "10 locations with atleast 30 yrs. of timeseries for each 20 sp.",
                                               "10 locations with atleast 30 yrs. of timeseries for each 20 sp.",
                                               "14 locations with atleast 45 yrs. of timeseries for each 22 taxa",
                                               "13 locations with atleast 50 dates of data")
tab_data_info$References<-c("US Rapid Carbon Assessment database (RaCA), Wills et al. (2014)", "McNab (2009)","McNab (2008)",
                            "Cedar Creek Ecosystem Science Reserve, biodiversity experiment e120",
                            "Rothamsted Insect Survey","Rothamsted Insect Survey","Continuous Plankton Recorder Survey",
                            "Great Miami Wetland Mitigation Bank, Smyth et al. (2019)")

knitr::kable(tab_data_info, 
             format="latex", align="l",
             caption = "Summary table for the data we used. Bolds entries are the multivariate data, the rest are bivariate datasets (see Data section). Basal metabolic rate=BMR \\label{tab_data_info}", 
             booktabs = T,linesep = "\\addlinespace") %>%
             column_spec(1:4,width="2.8 cm")%>%
             row_spec(5:8, bold=T)
```

@@

The third mechanism we explored is evolutionary, and pertains to bivariate 
trait data across species, e.g., to our bird and mammal BMR versus body mass 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million different
bivariate draws from distributions showing one of five distinct types of copula 
structures: 1) extreme left-tail dependence, 2) moderate left-tail dependence,
3) symmetric tail dependence, 4) moderate right-tail dependence, and 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech} for details). 
All marginal distributions were standard normal, and distributions all had Spearman 
correlation $0.875$ between their two components, so our simulations assess 
the impact only of copula structure in evolutionary character
changes on evolutionary outcomes. For each of the five copula structures, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across the phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

<!--Holder stuff-->
Our simulations of character evolution suggest the hypothesis that evolutionary
changes through time in bird and mammal BMR and body size may exhibit right-tail dependence. This is
a hypothesis only, because the right-tail dependence of Fig. \ref{fig_biv_multi_raw_cop}F,G
could have come about in another, unknown way: our simulations show that tail dependence 
in evolutionary changes is sufficient, but may not be necessary, to produce tail
dependence in characters of extant species. For instance, as discussed below and in
Appendix \ref{SM-sect:missingdata} (see also Fig. \ref{SM-fig:missingdata}), 
systematically missing data can produce tail dependence and may
have influenced results for the BMR versus body mass datasets. Even if the hypothesized 
evolutionary mechanism
is correct, our results simply replace one question, i.e., why do we see right-tail
dependence between BMR and body mass for birds and mammals, with another question,
i.e., why might we see right-tail dependence in evolutionary changes of these traits?
See Appendix \ref{SM-evol_mech} for additional thoughts on the potential importance of 
copulas for studies of character evolution when phylogeny branch lengths are not available.

<!--Discuss stuff here on tradeoff and life history that Bever told me about-->
The relationships we studied between BMR and body mass relate
to a trade-off between mass-specific BMR (i.e., BMR per unit body mass) and body
mass itself. Trade-offs between life history characteristics are a focal topic in 
evolutionary ecology. There are reasons to believe copulas interrelate with 
life-history theory in additional ways beyond what we could demonstrate 
with the data we used.
It is well known that energy allocation to a life function, X (e.g., reproduction)
will reduce the energy that can be allocated to other functions, $Y_1$, $Y_2$, $Y_3$ 
(e.g., growth, predation avoidance; this is sometimes called the principle of 
allocation).
But it is equally well known that X can trade-off against any or all of 
the $Y_i$. Therefore, for large values of X which approach absolute 
limitations, there may be a strong 
association between X and $Y_1$. Whereas for small values of X there may be
little association - resources not allocated to X can instead be allocated 
to $Y_1$ or $Y_2$ or to some combination. This constitutes 
asymmetric tail dependence between X and $Y_1$. For instance, \cite{Winemiller1992}
described a three-way trade-off in North American fishes between age of reproductive
maturity, juvenile survivorship, and fecundity. The trade-off should
produce a tight association between age of maturity and fecundity for 
fishes with low age of maturity, but little association between those
variables for later-maturing
fishes because those fishes may invest much or little in juvenile survival.
Although copulas may be very useful for studying multi-dimensional trade-offs, 
applications will 
require careful attention to the possible consequences of biased sampling:
if the degree of completeness of a dataset is associated with one or
more of the quantities being studied then statistical artefacts
may bias conclusions of copula analyses (Appendix \ref{SM-sect:missingdata}, 
Fig. \ref{SM-fig:missingdata}).

<!-- We have only given a few mechanisms, but there are likely to be others, and discoverig
additional mechanisms of copula structure is certainly worthy of future work. -->
Other mechanisms that we have not considered probably also operate to produce copula structure
in ecological and evolutionary datasets, and should be considered in future work. For instance, 
measurement error may modify copula structure, though one would have to have information
on the precise nature of measurement error in a dataset to explore its influence. 
Not only did the models we used to explore mechanisms not consider measurement error, 
they were intentionally simple in other respects, as well, not including phenomena such as 
delayed density dependence, dispersal, stage structure in population dynamics, and trophic
interactions.
We hope by enumerating a few potential mechanisms of copula structure in data 
we help inspire additional research on the
potentially numerous mechanisms that may operate in diverse datasets.

@@

We thank the many contributors to the large datasets we used; D. Stevens and P. Verrier for data 
extraction; and Joel E. Cohen, David Tilman, Lauren Hallet, Jonathan Walter, Thomas 
Anderson, and Lei Zhao for helpful suggestions. We thank James Bell of the Rothamsted Insect Survey 
(RIS). 
The RIS, a UK Capability, is funded by the Biotechnology and Biological Sciences Research Council 
under the Core Capability Grant BBS/E/C/000J0200. SG, LWS and DCR were partly funded by US National 
Science Foundation grants 0075516 and 1442595 and the James S McDonnell Foundation.
